---
title: "Review of Introductory Statistics"
author: "Zack Treisman"
date: "Spring 2021"
output: 
  beamer_presentation: 
    keep_tex: yes
    includes: 
      in_header: ../slide_style.tex
bibliography: ../bibliography.bib    
citecolor: blue
---

```{r setup, include=FALSE}

library(MASS)

```

## Data

An **observation** is a single unit.  A **variable** is a measurement made on that unit 

* Record observations as **rows** and variables in **columns**.    
* Variables can be **categorical** or **numerical**.
    * Categorical variables can be **binary** or not, **ordered** or not.
    * Numerical variables can be **discrete** or **continuous**.
* Dates, times and locations merit special consideration.
* Vocabulary is not universal: Factor, case, treatment \ldots

```{r}
head(Sitka) # from package MASS
```


## Distributions

The **distribution** of a variable is a measure of how often it takes each possible value.

* The distribution of a categorical variable is a list of the percentage of observations in each category.
```{r echo=TRUE}
table(Sitka$treat)/length(Sitka$treat)
```

## Distributions

* Picture the distribution of a numerical variable with a histogram, boxplot or density estimate.
```{r out.height="0.5\\textheight" }
hist(Sitka$size)
```

* Shape: center, spread, skew, kurtosis

##  Numerical summaries

```{r echo=TRUE}
summary(Sitka$size)
```

```{r}
quantile(Sitka$size, c(0.025,0.975))
```

```{r}
sd(Sitka$size)
```

```{r}
var(Sitka$size)
```

```{r}
IQR(Sitka$size)
```

## Common distributions

* Normal distribution
    * The sum of many independent effects tends to be normal.
    * Formula that you never use: $\displaystyle N(\mu,\sigma^2)(x)=\frac{1}{\sigma\sqrt{2 \pi}}e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$
    * Observations on disparate scales can be standardized with z scores: $\displaystyle z=\frac{x-\mu}{\sigma}$
    
\medskip    
    
* Other distributions from Intro Stats
    * t - Like the normal distribution, but adjusted for small samples.
    * $\chi^2$ - Sum of squared standard normals.
    * F - Similar to $\chi^2$, used in ANOVA.
    * Binomial - How many successes in $n$ trials?
    * Poisson - Count of discrete events in fixed time or space.
    * \ldots

## Inference

* Confidence intervals
* Hypothesis tests
    * p(robability)-values
    * Null and alternate hypotheses
    * t-tests, ANOVA, $\chi^2$ tests

\footnotesize   
```{r}
t.test(size~treat, data = Sitka)
```
\normalsize

## Always plot your data!

```{r}
boxplot(size~treat, data = Sitka)
```

## Linear models

* Slope and intercept parameters
* Correlation
* Residuals
* Inference

\scriptsize   
```{r}
summary(lm(size~Time, data = Sitka))
```
\normalsize


## Always plot your data!

```{r out.height="0.7\\textheight"}
plot(size~Time, data = Sitka)
abline(lm(size~Time, data=Sitka))
```

