---
title: "Review questions solutions"
author: "Zack Treisman"
date: "Spring 2021"
output: html_document
---

# 1 

There are many possible ways to answer this question. Here is one.

(a) i: Cases are observations of individual sheep at a particular place and time, ii: possible variables are sex, age, location

(b) Sex and location are categorical, though location might be usefully numerically coded, say with lat and long. Age could be numerical if we have a way of knowing, or it could be an ordered categorical variable (for example: young, adult, old).

(c) Do male and female sheep frequent the same areas?

(d) Sex is explanatory, location is response.

# 4


(a) What proportion of the fruit flies have thorax length less than 0.72 mm?
```{r}
pnorm(0.72,0.8,0.078)
```

(b) What proportion of the fruit flies have thorax length greater than 0.82 mm?
```{r}
pnorm(0.82,0.8,0.078, lower.tail = FALSE)
```

(c) What proportion of the fruit flies have thorax length between 0.7 and 0.9 mm?
```{r}
pnorm(0.9,0.8,0.078) - pnorm(0.7,0.8,0.078)
```

(d) We wish to select the fruit flies with the highest 20% of thorax length. What is
the shortest thorax length we should consider?
```{r}
qnorm(0.8,0.8,0.078)
```

# 5

(a) Which of these five variables are numerical? income and age

(b) Which type of plot would be appopriate to use to view the relationship between
income and gender? hollow histograms or side by side boxplots

(c) All but iii are true.

(d) Only iii is true.

# 8 

(a) What is your point estimate for the proportion of Western students who would
consider a physics major who are female?
```{r}
ptEst <- 23/50
```

(b) Using a normal model for the proportion, what is the standard error in your
estimate?
```{r}
se <- sqrt(23/50*27/50/50)
se
```

(c) Give a 95% confidence interval for the proportion of potential physics majors at
Western who are female.
```{r}
prop.test(23,50)
```
alternatively
```{r}
lower <- ptEst - 2*se
upper <- ptEst + 2*se
conf.interval <- c(lower,upper)
conf.interval
```

(d) If you would like your margin of error to be at most ±5%, assuming that the
proportion of potential physics majors who are female does not change, how
many potential physics majors would you have to include in your sample?
```{r}
(23/50)*(27/50)/(.05/1.96)^2
```

# 12

(a) What are your null and alternate hypotheses? $H_0$: The distributions of fish are the same in both lakes. $H_a$: The lakes do not have the same distribuitons of fish.

(b) How many of fish A do we expect to see in Green Lake if the distributions are
the same in both lakes?

```{r}
1182*1058/2045
```

(c) What are the degrees of freedom, the $\chi^2$ statistic, and the p-value?

```{r}
fish<-matrix(c(650, 412, 120, 408, 387,68), ncol=2)
chisq.test(fish)
```

We can also answer (b) using:
```{r}
chisq.test(fish)$expected[1,2]
```

(d) Do you think that the distributions of fish types are the same in both lakes?
Explain. The p-value is very small, it seems unlikely that the true distributions are the same in both lakes.

# 13

(a) Give appropriate null and alternative hypotheses.
$H_0: \mu=1500$, $H_a: \mu\neq1500$

(b) Give the test statistic, degrees of freedom, and p-value for the test.
(c) Give a 95% confidence interval for the mean metabolic rate in this group.

```{r}
rates<-c(995, 1425, 1396, 1418, 1502, 1256, 1189, 913, 1124, 1052, 1347, 1204)
t.test(rates, mu=1500)
```

(d) Clearly interpret your results in a sentence. The p-value is small, it seems unlikely that the mean metabolic rate in this population from which this sample was taken is 1500 kCal/day.

# 16

(c) is the correct calculation.

# 17

(a) Do you trust the regression line? Why or why not? Sure. The residuals show no clear pattern, and the correlation coefficient is reasonably large.

(b) What is the biological meaning of the intercept $395.01$? Do you trust this? Why
or why not? This would be the area of a wing with zero weight. Not a meaningful extrapolation.

(c) What is the biological meaning of the slope $38.23$? This is the expected increase in wing area per unit increase in wing weight, or the average area of one mg of butterfly wing.

(d) What does the model predict for the area of a wing weighing 10 milligrams? $395.01+38.23 \times 10=777.31$