---
title: "Mixed models"
output: 
  html_document:
    highlight: tango
    theme: readable
    toc: true
    toc_float: true
bibliography: "../bibliography.bib"   
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
knitr::opts_knit$set(global.par = TRUE)
```

```{r message=FALSE}
library(car)
library(glmmTMB)
```


## Overview

Your array of modeling tools is now quite extensive. This lab sets you to work on reproducing the results of some complicated analysis.

## Setting

The paper you will analyze is @salamanders. It is available from the [Journal of Applied Ecology](https://besjournals.onlinelibrary.wiley.com/doi/epdf/10.1111/1365-2664.12585). The data are in the `glmmTMB` package as `Salamanders` and are available from the [Dryad data repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.5m8f6). For a data dictionary and other information about the data see `?Salamanders`.

## Linear models

We can begin by asking the data a very simple question: "Were salamanders more abundant in areas that were not mined?" We'll use `glmmTMB` even though all we are doing is making a linear model with Gaussian errors (one-way ANOVA) so that when we compare models, all of them will have been built with the same tool.

```{r}
Msal1 <- glmmTMB(count~mined, data=Salamanders)
summary(Msal1)
```

(@) What does this model tell us?

We can add in other predictors from the data.

```{r}
Msal2 <- glmmTMB(count~mined+cover+DOP+Wtemp+DOY+spp, data = Salamanders)
Anova(Msal2)
```
(@) Based on which variables appear significant in this model, make a graphic or graphics showing the relationships between some of these predictors and the response.

## Extending to a GLM

Since the response is a count, it might be better to model this with a Poisson or negative binomial GLM. In `glmmTMB`, there are two negative binomial families. What most others call negative binomial is `negbinom2`, and `negbinom1` is often called quasi-poisson.

```{r}
Msal3 <- glmmTMB(count~mined+cover+DOP+Wtemp+DOY+spp, data = Salamanders, 
                 family = "poisson")
Msal4 <- glmmTMB(count~mined+cover+DOP+Wtemp+DOY+spp, data = Salamanders, 
                 family = "nbinom1")
Msal5 <- glmmTMB(count~mined+cover+DOP+Wtemp+DOY+spp, data = Salamanders, 
                 family = "nbinom2")
```

(@) Decide which distribution family seems most appropriate, and justify your decision.

## Adding a random effect

The data contains repeated samples of the same site, so there is pseudoreplication in the models we have built so far. This can be accounted for by including `site` as a random effect.

```{r}
Msal6 <- glmmTMB(count~mined+cover+DOP+Wtemp+DOY+spp + (1|site), 
                 data = Salamanders, 
                 family = "nbinom1") # Match this to your choice above.
```

(@) Report the results of including the random effect, and how this changed the model.

## Zero inflation

In the paper, they describe modeling occupancy distinctly from abundance. We can include step this via a zero inflation term. We'll follow the authors of the study and include only the environmental covariates `cover`, `DOP`, `Wtemp` and `DOY` in the occupancy part of the model.

```{r}
Msal7 <- glmmTMB(count~mined+cover+DOP+Wtemp+DOY+spp + (1|site), 
                 data = Salamanders, 
                 zi = ~cover+DOP+Wtemp+DOY,
                 family = "nbinom1") # Again, match this to your choice above.
```

## Finding an optimal model and matching the results of the paper

In the paper, they consider the interaction between mining and species `mined*spp` for predicting counts. Additionally, they include the environmental covariates `cover`, `DOP`, `Wtemp` and `DOY` only in the occupancy part of the model, not the abundance part.

(@) Find your optimal model, using all of the ingredients discussed above. Do your results agree with those in the paper? (A Bayesian procedure was used to produce the published results.)

## References