---
title: "Population Assessment and Distance Sampling"
author: "Zack Treisman"
date: "Spring 2021"
output: 
  beamer_presentation: 
    includes: 
      in_header: ../slide_style.tex
bibliography: ../bibliography.bib    
citecolor: blue
---

```{r setup, include=FALSE}
knitr::opts_knit$set(global.par = TRUE)
library(ggplot2)
library(Distance)
set.seed(33)
```

```{r,include=FALSE}
par(mar = c(4, 4, 0.5, 0.5)) # Set margins
scale_colour_continuous <- scale_colour_viridis_c
scale_colour_discrete   <- scale_colour_brewer(palette="Dark2")
scale_colour_binned     <- scale_colour_viridis_b
``` 

## Philosophy

Wildlife population assessment -  *How many are there?* 

## Simpler questions:

* rank
* occupancy
* index

## Population

* census
* mark-recapture
* removal
* plot sampling
* distance sampling

## A simulated example

We'll use some of the code from the section on simulation to generate random arrangements of populations.

```{r, include=FALSE}
random_pop <- function(Lx=30, Ly=30,
                         nparents=50, offspr_per_parent = 10, dispdist=2){

  parent_x = runif(nparents,min=0,max=Lx) 
  parent_y = runif(nparents,min=0,max=Ly)
  noffspr = nparents*offspr_per_parent
  angle = runif(noffspr,min=0,max=2*pi) 
  dist = rexp(noffspr,1/dispdist)

  offspr_x = rep(parent_x,each=offspr_per_parent)+cos(angle)*dist
  offspr_y = rep(parent_y,each=offspr_per_parent)+sin(angle)*dist
  pos <- as.data.frame(cbind(offspr_x,offspr_y))[(offspr_y>0)&(offspr_y<Ly)&(offspr_x>0)&(offspr_x<Lx),]

  return(pos)
}
```

\scriptsize
```{r}
pos <- random_pop()
```
\normalsize
The simulated population size is `r nrow(pos)`.
\scriptsize
```{r  fig.height = 3, fig.width = 6}
ggplot(data=pos, aes(offspr_x, offspr_y))+geom_point()+labs(x="", y="")
```

## Capture-mark-recapture

* Capture a subset of the population and mark them. 
* Release. After sufficient time for the released subset to have fully mixed with the population, capture another subset. 
* The fraction of the second capture that is marked estimates the fraction of the population in the first capture. 
* As a formula: $\hat P = C_1/p_m$ where 
    * $\hat P$ is the population estimate, 
    * $C_1$ is the size of the first captured group and
    * $p_m$ is the marked proportion in the second captured group.

\scriptsize
```{r}
capture1 <- sample(nrow(pos), 50) 
pos$mark <- 1:nrow(pos) %in% capture1

capture2 <-  sample(nrow(pos), 50)
pos$recapture <- (1:nrow(pos) %in% capture2) & pos$mark
```
\normalsize

* There are `r sum(pos$recapture)` recaptures. 
* This is  `r sum(pos$recapture)/length(capture2)` of the second captured group.
* The population estimate is `r length(capture1)`/`r (sum(pos$recapture)/length(capture2))`=`r round(length(capture1)/(sum(pos$recapture)/length(capture2)))`.

## Plot of captures and recaptures

This method is not particularly well suited to populations that don't move around and mingle.

\scriptsize
```{r  fig.height = 3, fig.width = 6}
ggplot(data=pos, aes(offspr_x, offspr_y, color=mark, shape=recapture))+
  geom_point()+labs(x="", y="")
```


## Capture-mark-recapture is fundamentally Bayesian

Population estimates can be improved over time by repeating the experiment. 

## Plot Sampling

Even if the individuals being counted don't move around, it is often impractical/unreasonable/impossible to census an entire area of interest. Instead, we often census appropriately chosen subplots.

* Subplots are often laid out along transects.

```{r  echo=F, fig.height = 3, fig.width = 6}
pos$in_plot <- ((pos$offspr_x+2.5)%%5<1)|((pos$offspr_x+2.5)%%5>4)
ggplot(data=pos, aes(offspr_x, offspr_y, color=in_plot))+geom_point()+labs(x="", y="")+
  geom_vline(xintercept=(5*(0:5)+2.5))+
  geom_vline(xintercept=(5*(0:5)+1.5), lty=2)+
  geom_vline(xintercept=(5*(0:5)+3.5), lty=2) + 
  theme(legend.position = "none") 
```

## Computing population from plot sampling

Divide the number counted in the plots by the fraction of the area that is covered by the plots.

\scriptsize
```{r}
num_plots <- 5
half_width <- 1
Lx <- 30; Ly <- 30
counted <- sum(pos$in_plot)
area_plots <- Ly*2*half_width*num_plots
total_area <- Lx*Ly
area_fraction <- area_plots/total_area
(pop_estimate <- counted/area_fraction)
```
\normalsize
We can use a bootstrap to get a confidence interval for this estimate.

## Distance sampling

What if detecting the individuals in the population is challenging?

* In many surveys, the further the individual is from the observer, the lower the probability of observation.

\scriptsize
```{r, include=FALSE}
pos$d <- NA
pos$d[(pos$offspr_x+2.5)%%5<1] <- (pos$offspr_x+2.5)[(pos$offspr_x+2.5)%%5<1]%%5
pos$d[(pos$offspr_x+2.5)%%5>4] <- 5-(pos$offspr_x+2.5)[(pos$offspr_x+2.5)%%5>4]%%5

pos$prob_detected <- 0
pos$prob_detected[(pos$offspr_x+2.5)%%5<1] <- dnorm((pos$offspr_x[(pos$offspr_x+2.5)%%5<1]+2.5)%%5,sd=0.5)/dnorm(0, sd=0.5)
pos$prob_detected[(pos$offspr_x+2.5)%%5>4] <- dnorm(5-(pos$offspr_x[(pos$offspr_x+2.5)%%5>4]+2.5)%%5,sd=0.5)/dnorm(0, sd=0.5)

pos$detected <- as.logical(rbinom(nrow(pos),1,pos$prob_detected))

```


```{r  echo=F, fig.height = 3, fig.width = 6}
ggplot(data=pos, aes(offspr_x, offspr_y, color=detected))+geom_point()+labs(x="", y="")+
  geom_vline(xintercept=(5*(0:5)+2.5))+
  geom_vline(xintercept=(5*(0:5)+1.5), lty=2)+
  geom_vline(xintercept=(5*(0:5)+3.5), lty=2) + 
  theme(legend.position = "none") 
```

```{r}
counted <- sum(pos$detected)
(pop_estimate <- counted/area_fraction)
```
## Overall detection probability

It is useful to look at a histogram of the distance of detected individuals.

* Goal: fit a detection probability function to these data.

```{r echo=F, fig.height = 3, fig.width = 6}
prob_detect <- function(x) dnorm(x,sd=0.5)/dnorm(0, sd=0.5)
ggplot(pos[pos$detected,], aes(d))+geom_histogram(breaks=(0:10)/10, aes(y = 6 * stat(count) / sum(count)))+labs(x="Distance from Centerline", y="Probability of Detection")+geom_function(fun=prob_detect, aes(color="red")) + 
  theme(legend.position = "none") 
overall_prob <- integrate(prob_detect, lower = 0, upper = 1)$value # in general, divide by w
```
The overall probability of detection is $\hat P=$ `r round(overall_prob, 4)`.

## Estimation using $\hat P$

To get an estimated count from a count where the probability of detection is not 100%, divide the actual count by the overall probability of detection. 

\scriptsize
```{r}
est_count <- counted/overall_prob
(pop_estimate <- est_count/area_fraction)
```
## Guessing the detection probability function

We have some data that we would like to use to define a curve.

* Propose a shape for the detection probability function (half-normal, hazard rate, uniform, \ldots).
    * $p_{det}(0)=1$ (objects at the centerline are detected)
    * $p'_{det}(0)=0$ (detection probability is rather flat around 0)
    * $p'_{det}(x)\leq0$ (detection probability decreases with distance)
    
* Use maximum likelihood to find parameters of the most appropriate curve in the family.

The $\hat P$ above comes from rescaling a normal distribution with $\sigma=0.5$ so that $p_{det}(0)=1$. The choice of $\sigma$ was from fitting by eye. The code below finds the $\sigma$ for the maximum likelihood. 
\scriptsize
```{r, warning=FALSE}
samp_ind<-pos[pos$in_plot,]
prob_detect2 <- function(x,s) dnorm(x,sd=s)/dnorm(0, sd=s)
nll_fn<-function(s) -sum(dbinom(samp_ind$detected,
                                1,prob_detect2(samp_ind$d,s),
                                log = TRUE))
opt1 <- optim(fn = nll_fn, par = list(s=0.5), method="BFGS"); opt1$par
```
## Using the optimal half-normal detection probability

```{r echo=F, fig.height = 3, fig.width = 6}
ggplot(pos[pos$detected,], aes(d))+geom_histogram(breaks=(0:10)/10, aes(y = 6 * stat(count) / sum(count)))+labs(x="Distance from Centerline", y="Probability of Detection")+geom_function(fun=prob_detect2, aes(color="red"), args = c(s=0.434)) + 
  theme(legend.position = "none") 
```

\scriptsize
```{r}
overall_prob2 <- integrate(function(x) prob_detect2(x,s=opt1$par),0,1)$value 
est_count2 <- counted/overall_prob2
(pop_estimate2 <- est_count2/area_fraction)
```

## Hazard rate detection probaility functions.

A half-normal detection probability function may not be best.

The hazard rate detection probability is defined by
$$
p_{det}(x)=1-\exp(-(x/\sigma)^{-b}).
$$
This function has two parameters, $\sigma$ which controls the width of the shoulder, and $b$ which controls the steepness of the dropoff.

```{r echo=F, fig.height = 3, fig.width = 6}
prob_detect_hr <- function(x,s,b) 1-exp(-(x/s)^(-b))
ggplot(pos[pos$detected,], aes(d))+geom_histogram(breaks=(0:10)/10, aes(y = 6 * stat(count) / sum(count)))+labs(x="Distance from Centerline", y="Probability of Detection")+geom_function(fun=prob_detect_hr, aes(color="red"), args = c(s=0.5,b=4)) + 
  theme(legend.position = "none") 
```

## Maximum likelihood to find the optimal hazard rate

Again, we can use maximum likelihood to estimate the parameters.

\scriptsize
```{r, warning=FALSE}
prob_detect_hr <- function(x,s,b) 1-exp(-(x/s)^(-b))
nll_fn_hr <- function(pars){s<-pars[1];b<-pars[2]
-sum(dbinom(samp_ind$detected,1,prob_detect_hr(samp_ind$d,s,b), log = TRUE))}
opt2 <- optim(fn = nll_fn_hr, par = list(s=0.25, b=4)); opt2$par
```

## The optimal Hazard Rate detection curve

```{r echo=F, fig.height = 3, fig.width = 6}
ggplot(pos[pos$detected,], aes(d))+geom_histogram(breaks=(0:10)/10, aes(y = 6 * stat(count) / sum(count)))+labs(x="Distance from Centerline", y="Probability of Detection")+geom_function(fun=prob_detect_hr, aes(color="red"), args = opt2$par) + 
  theme(legend.position = "none") 
```

## Using the optimal Hazard Rate function to predict $\hat P$

And then we estimate 

* the overall probability $\hat P$ by integrating the detection probability function, 
* the number of individuals within the sampling region by dividing the count by $\hat P$, and 
* the population by dividing the number of individuals by the area fraction.

\scriptsize
```{r}
overall_prob3<-integrate(function(x){
  prob_detect_hr(x,s=opt2$par[1],b=opt2$par[2])},0,1)$value  
est_count3 <- counted/overall_prob3
(pop_estimate3 <- est_count3/area_fraction)
```

## Assumptions and issues

* If the individuals that are being counted are clustered (pods of whales, clusters of plants, \ldots), this can cause problems.

    * Estimate the distance to the center of the cluster.
    * Record the size of each cluster and estimate the mean cluster size.

* Often, instead of using all observations, it makes sense to truncate observations above some effecting strip width.

* If there is avoidance of the observer, this can cause problems.

## Point-transect distance sampling

For studies where the observer is stationary, the process is quite similar.
 

## References
