---
title: "Linear Regression"
output:
  pdf_document:
    highlight: tango
colorlinks: true
linkcolor: blue    # internal links (sections, refs)
urlcolor: blue     # web URLs
citecolor: blue    # bibliography citations
bibliography: ../bibliography.bib
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

library(ggplot2)

library(car)

```

# Overview

The goal of this lab is to give you practice using R to create and evaluate basic regression models. This includes both classical linear regression and analysis of variance. Even though there are certainly differences between these two scenarios, seeing them as aspects of a common technique can make life easier.

We'll be using `ggplot2` to make graphics and `Anova` from the `car` package in this lab so load those libraries.

```{r, eval=FALSE}

library(ggplot2)

library(car)

```

# Data requirements and options

In this lab you will work with one dataset, which may be either the redside shiner data provided below or another dataset of your choice that you have domain knowledge about.

To make sure the same analyses make sense for everyone, your dataset (whether shiners or your own) should have at least:

- at least one **quantitative response** variable (e.g., mass, test score, time, outcome);
- at least one **categorical explanatory** variable with 3–5 levels (e.g., location, treatment group, school, position);
- at least one **quantitative explanatory** variable for simple regression (e.g., length, baseline score, age);
- optionally, a second categorical or quantitative explanatory variable so that we can build interaction models.

For the purposes of writing generic code, it can be helpful to think in terms of the following roles:

- `my_data`: your main data frame;
- `y_continuous`: a quantitative response variable;
- `group_factor`: a categorical variable with 3–5 levels for one-way ANOVA;
- `x_continuous`: a quantitative explanatory variable for simple regression;
- `factor2`: a second factor (optional; used later for interaction models).

If you are using your own dataset instead of the shiner data, you should identify variables that play these roles and, if you wish, define a small mapping at the top of your script, for example:

```{r, eval=FALSE}
# Example mapping (modify if you use your own data)
my_data      <- shiners
y_continuous <- "smass"
group_factor <- "loc"
x_continuous <- "ssl"
factor2       <- "temp"
```

Later in the lab, whenever you see code that uses the shiner variables directly (for example, `smass`, `loc`, `temp`), you may substitute your own variables that serve the same purpose in your chosen dataset.

# The data: Growth of redside shiners

The data that we will be looking at today are from @houston. The goal of the experiment was to determine whether observed differences in fish growth resulted from environmental or genetic variation.

If you would like to work with a different topic that you know something about, you may replace the redside shiner data with another data set that has at least one quantitative response variable and at least one categorical explanatory variable (with 3–5 levels) so that the analyses in this lab still make sense. To find candidate data sets, you can use the AI prompt: “I am working on a regression and ANOVA lab in R. I need three public data sets I can download as CSV files from the internet. Each data set should match a topic I am interested in (for example sports, education, environment, health, music, or social media), should have at least one quantitative response variable and at least one categorical explanatory variable with 3–5 levels, and should be suitable for fitting linear models and one-way ANOVA using lm() and Anova() in R. For each of three different topics of my choosing, suggest one specific data set, give a one-sentence description, and provide a direct download link to the CSV file.” or you may search on your own (for example, on Kaggle or data repositories) for a dataset you understand.

The redside shiner data are on GitHub (https://raw.githubusercontent.com/ztreisman/Statistical_Modeling/main/03Regression_and_ANOVA/data/redside_shiner.csv) or on [OneDrive](https://westernstatecoloradou-my.sharepoint.com/:x:/g/personal/ztreisman_western_edu/IQCoC-SG8ZH3SpEy4GG2TT8SATBoHdlgz4e1UpP0vX3supk?e=gmLuEa) in the file *redside_shiner.csv*. Save this file to the data folder in your working directory for this lab and load it as the data frame `shiners` with the command

```{r, eval=TRUE}

shiners <- read.csv("data/redside_shiner.csv")

```

Run `str(shiners)` or click on the blue circle with the white triangle icon next to the data in the *Environment* pane to get a list of the variables and their types. Observations in this data set are individual fish. The variables are:

* `obs` is the observation number

* `loc` is location

* `block` is a part of the experimental design

* `temp` is temperature where the fish were grown in deg C, an experimental treatment

* `food` is frequency of feeding, another experimental treatment

* `smass` is the starting mass

* `emass` is the mass at the end of the experiment

* `ssl` is the starting standard length

* `esl` is the ending standard length

* `days` is the length of time that the fish was in the experiment

Use `summary(shiners)` and `View(shiners)` to get an idea of the data. There are a fair number of observations with missing data, but the data set is still sufficiently large if we drop the observations with missing entries, and it makes the analysis much easier.

```{r}

shiners <- na.omit(shiners)

```

Since `temp` only takes three values, it makes sense to convert it to a factor.

```{r, eval=TRUE}

shiners$temp <- factor(shiners$temp)

```

If you are using your own dataset, identify and document the variables that will play the roles of:

- quantitative response (analogous to `smass` or `emass`),
- categorical factor for ANOVA (analogous to `loc`),
- quantitative predictor for regression (analogous to `ssl` or `esl`),
- and optional additional factors or covariates (analogous to `temp`, `food`, or `smass` as a covariate in the growth models).

# Starting populations

In this section we look at the beginning of the experiment and explore whether groups differ at baseline. With the shiner data, we use starting length and mass by location; with other datasets, you should choose analogous variables.

For the redside shiner data, let's examine the fish as they were at the beginning of the experiment. Make a plot showing `smass`, `ssl` and `loc`.

```{r fig.width=7}
ggplot(shiners, aes(ssl, smass, color = loc))+
geom_point()+
labs(x = "Starting Standard Length (cm)", y = "Starting Mass (g)",
title = "Starting Fish Sizes", color = "Location")
```

(@) If you are using the shiner data: Does it appear that the fish from any one location are larger or smaller than the others either in mass, length or both?

(@) If you are using your own data: Make an analogous plot with a quantitative response and a quantitative explanatory variable, colored by a categorical grouping variable (e.g., score vs baseline, grouped by region)


We can do a one-way anova to answer this question for mass (or another continuous response) across groups. Start by examining the normality and homogeneity of variance of the response for the groups to confirm that ANOVA is legitimate.

For the shiner data:

```{r fig.width=7}
ggplot(shiners, aes(smass))+
geom_density()+
facet_wrap(~loc)
ggplot(shiners, aes(loc, smass))+
stat_summary(fun=var, geom = "bar")+
labs(x="Location",y="Variance in Starting Mass")

tapply(shiners$smass, shiners$loc, var)
```

If you are using your own dataset, replace `smass` with your chosen quantitative response and `loc` with your grouping variable.

We perform the regression with the `lm` function.

```{r}
lm1 <- lm(smass~loc, data = shiners)

summary(lm1)
```

If you are using your own data, fit an analogous model `lm(response ~ group_factor, data = my_data)` and examine the output.

The last line of the output gives us the p value for the \(F\) test (\(p<0.001\)), and we can conclude that there is some difference in starting mass. The coefficients of the model tell us that while the Beaver Creek and Snake River fish do not appear different, even after correcting the p value for multiple comparisons (there are \({3\choose2}=\frac{3!}{2!(3-2)!}=3\) comparisons, so multiplying the coefficient of `locMainCreek`'s p value of \(0.000238\) by 3 gives the Bonferroni correction, which tends to be conservative) we have \(p<0.001\) for the comparison of Main Creek and Beaver Creek. Mean starting mass for the Main Creek fish appears to be about \(0.1\)g greater than the mean masses for the fish from the other locations. With only three levels to the single factor in this model it is possible to see the individual comparisons in the coefficients table, but in general after a significant \(F\) test we will want to perform post-hoc tests with R.

```{r}
pairwise.t.test(shiners$smass, shiners$loc, p.adjust.method = "bonferroni")
```

(@) For the shiner data: Are the Main Creek fish also longer on average at the start of the experiment?

(@) For your own dataset: Perform a similar one-way ANOVA on a different but related response (for example, length instead of mass, math score instead of reading). Do the conclusions about differences among groups change?

## An aside on dimensions and transformations

Length and mass are both measuring size. Since mass is proportional to volume, and volume is measured in cubic cm while length is measured in cm, it makes sense to expect that the relationship between mass and length is roughly cubic.

\[
\text{mass}=k\cdot\text{length}^3
\]

where \(k\) encompasses density and aspect ratio. Significant deviations from this exponent of 3 could potentially have interpretations in terms of the growth processes of the fish but we won't speculate on that here. Taking the log of this equation gives

\[
\log(\text{mass})=\log(k)+3\log(\text{length})
\]

Plotting this relationship can either be done by transforming the variables, or using a log scale.

For the shiner data, let's examine the fish as they were at the beginning of the experiment. Make a plot showing `smass`, `ssl` and `loc`, on log scales.

```{r}
ggplot(shiners, aes(ssl, smass, color = loc))+
geom_point()+
labs(x = "Starting Standard Length (cm)", y = "Starting Mass (g)",
title = "Starting Fish Sizes (log scale)", color="Location")+
scale_x_log10() + scale_y_log10()
```

This does look like a more linear pattern than the plot made on the direct scales above. We can use a regression to confirm this relationship.

```{r}
lm2 <- lm(log(smass)~log(ssl), data = shiners)

summary(lm2)
```

To evaluate this regression, we can plot the residuals against the explanatory variable.

```{r}
plot(lm2$residuals~log(shiners$ssl))

abline(0,0, lty="dashed")
```

A linear model can also be evaluated by the four diagnostic plots produced by R. When you run the following line look at the console for a prompt to scroll through the four plots that follow.

```{r}
par(mfrow=c(2,2)) # to show all four plots at once

plot(lm2)

par(mfrow=c(1,1)) # to see one plot at a time again
```

We examine the first and third of these diagnostic plots for evidence of heteroscedatsicity. The second helps us assess normality of the residuals, and the fourth helps us assess the influence of any detected deviations from these assumptions. There do not appear to be any significant issues with the assumptions underlying this regression.

We might wonder if location has any connection on this relationship. It can be introduced to the regression additively, so that it can only change the intercept, which is \(\log(k)\):

```{r}
lm2a <- lm(log(smass)~log(ssl)+loc, data = shiners)

anova(lm2a)

summary(lm2a)
```

Or it can be an interaction term, so that it can potentially change the slope, which is in this case the exponent of approximately 3.

```{r}
lm2b <- lm(log(smass)~log(ssl)*loc, data = shiners)

anova(lm2b)
```

The same checks for normality and homogeneity of variance in `smass` across locations above are also relevant to this regression.

It appears that the differences in mass due to location that we noted earlier are also reflected in this regression, though only the additive effect is only significant at the \(\alpha=0.05\) level, not the interaction effect.

(@) For the shiner data: Does this result change after the experiment? Perform the same regressions for location's effect on the relationship between \(\log(\text{emass})\) in terms of \(\log(\text{esl})\). Report on your results, including the diagnostic checks.

(@) For your own data: Identify a pair of continuous variables where you believe a power or log relationship might make sense (for example, income vs population, area vs perimeter, speed vs distance). Fit a log–log regression and comment on the estimated slope and on the diagnostic plots.

# The Experiment

The goal is to evaluate growth, so it makes sense to create two new variables measuring the differences in mass and length at the beginning and end of the experiment. To create the mass difference variable for the shiner data:

```{r}
shiners$mass.diff <- shiners$emass - shiners$smass
```

(@) Also create a length difference variable. Call it `sl.diff`.

More generally, if your dataset has pre/post or before/after measures, you can create a change variable as

\[
\text{change} = \text{after} - \text{before}.
\]

(@) If you are using your own data, identify a pair of “before” and “after” measurements (for example, pre- and post-test scores, start and end of season performance, baseline and follow-up). Create an analogous change variable and describe what a positive change means in your context.

## From t tests to regression

Let's begin by asking if the mass difference is the same in the two feeding groups. We'd expect that the fish who were fed twice daily grew more, but why not test this?

First, check the distributions for these two groups. There appears to be some right skew but it isn't severe.

```{r}
ggplot(shiners, aes(mass.diff))+
geom_density()+
facet_wrap(~food)
```

To do a t test, we can use the `t.test` function:

```{r}
t.test(mass.diff~food, data = shiners)
```

Alternatively, we can try to answer the same question with regression:

```{r}
summary(lm(mass.diff~food, data = shiners))
```

Note that the p value is almost exactly the same. In fact, if you run `t.test` with the option `var.equal=TRUE`, you will get exactly the p value from the regression. The advantage of the t test is its ability to deal with unequal variances.

(@) For the shiner data: Do a t test or regression to determine if `food` has an effect on `sl.diff`. Confirm the suitability of the test and report the results.

(@) For your own data: Choose a binary factor (for example, intervention vs control, male vs female, home vs away). Perform a t test and then a regression of your change variable (or another continuous outcome) on this factor. Compare the p values and discuss when you might prefer one approach over the other.

## Analysis

The results of the experiment stated in @houston are:

* _Individuals grew faster at higher temperatures and with more food, and there was a significant interaction between location and temperature. There was no difference in growth rates among the three populations at 10C and 17C. However, at 24C, individuals from the Snake River population grew significantly slower than those from Badger Creek and Main Creek._

Let's come to this same conclusion. We'll fit a full interaction model. That's a lot of terms, but only a few of them are statistically significant. In many cases, especially observational studies, there is not enough data to test all of the interaction terms. When we have more terms in the model, we lose power. To test interactions only up to a specific level, you can use the symbol `^`. For example, to only test up to the two term interactions, replace `smass*block*loc*temp*food` with `(smass+block+loc+temp+food)^2` in the formula for `lm.mass`.

For the shiner data, we use:

```{r}
lm.mass <- lm(mass.diff ~ smass*block*loc*temp*food, data=shiners)

Anova(lm.mass)
```

Before we continue, we should make some diagnostic plots. We can start with the standard diagnostics from plotting `lm.mass`.

```{r}
par(mfrow=c(2,2))

plot(lm.mass)

par(mfrow=c(1,1))
```

There do appear to be some deviations from both heteroscedasticity (visible in the first plot) and normality (visible in the second plot). Let's also look at the residuals and the continuous variable, `smass`.

```{r}
plot(lm.mass$residuals~shiners$smass)

abline(h=0)
```

There doesn't seem to be an issue with this variable alone. To be thorough, we should look at all of the others. Instead we'll revisit these data when we have some more tools at our disposal, but for now let's continue with the present analysis.

In order to probe the effects of the significant factors, we can build a simpler model that only includes these terms marked as significant in the full model.

```{r}
lm.mass2 <- lm(mass.diff ~ loc+temp+food+loc:temp, data=shiners)

Anova(lm.mass2)

summary(lm.mass2)
```

As reported in the paper, it appears that the fish grew faster at higher temperatures and with more food (the coefficients on `temp17`, `temp24` and `foodTwice Daily` are significant and positive), and the individuals from the Snake River grew more slowly at 24C (the coefficient on `locSnake River:temp24` is significant and negative). We can see this last effect in the following plot.

```{r}
ggplot(shiners, aes(temp, mass.diff, color = loc, group = loc))+
stat_summary(fun.data = mean_cl_normal,
position = position_dodge(width = 0.2))+
stat_summary(fun = mean, geom = "line",
position = position_dodge(width = 0.2))+
labs(x = "Temperature (C)", y = "Gain in Mass (g)", color = "Location")
```

(@) Conduct the same analysis, but for `sl.diff`. Does it tell the same story?

(@) For students using their own data: Identify one continuous response, one continuous covariate, and two categorical factors in your dataset (or as many as you can reasonably justify). Fit a model with all main effects and selected interactions, analogous to the model above, and then fit a reduced model keeping only the terms that appear important. Use `Anova` and `summary` to interpret the main effects and interactions in the context of your domain (e.g., education, sports, environment). Write 2–3 sentences translating the statistically significant results into plain language for a domain expert.

## References